Sampling deaths and OLS

### 2018-01-21 ols.Rmd
go on

So the *e60* for men from Skellefteå is `r round(x$coef[1], 2)` and for women it is `r round(x$coef[2], 2)` more years.

# Sampling deaths

We repeat the analyses with the data subset consisting of all those who were observerd to die in the parishes.

```{r smallols}
ids <- ols$id[ols$event == 1]
sols <- ols[ols$id %in% ids, ]
```

## Ordinary *Cox regression*

```{r standard}
fit <- coxph(Surv(enter, exit, event) ~ sex + region, data = sols)
survfit(fit)
```

```{r aft}
fit <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "lognormal", param = "lifeExp")
meanlog <- fit$coefficients["log(scale)"]
sdlog <- 1 / exp(fit$coefficients["log(shape)"])
fit$baselineMean <- exp(meanlog + sdlog^2 / 2)
summary(fit)
```

```{r plotln}
plot(fit, fn = "haz", xlab = "Years after 60.")
```

```{r aftgomp}
fit.g <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "gompertz", param = "lifeExp")

```{r plotgomp}
plot(fit.g, fn = "haz", xlab = "Years after 60.")
```

# OLS regression

Finally, let us perform ordinary OLS regression.

```{r ordOLS}
ools <- sols[sols$event == 1, ]
summary(fit <- lm(exit ~ sex + region, data = ools)) 
```

We first try to get a picture of what really is happening in mortality relative to `sex` and `region` (the figure is truncated at duration 30, i.e., age 90).

>>>>

So the *e60* for men from Skellefteå is `r round(x$coef[1], 2)` years and for women it is `r round(x$coef[2], 2)` more years.

## Sampling the dead

We repeat the analyses with the data subset consisting of all those who were observerd to die in the parishes. This implies that 13 percent censored individuals are thrown away.

```{r smallols, echo = FALSE}
ids <- ols$id[ols$event == 1]
sols <- ols[ols$id %in% ids, ]
```

```{r survrnocens}
fit2 <- survreg(Surv(exit, event) ~ sex + region, data = sols, dist = "gaussian")
x2 <- summary(fit2)
x2
```

So the *e60* for men from Skellefteå is now `r round(x2$coef[1], 2)` years and for women it is `r round(x2$coef[2], 2)` more years.

## Conclusion

Sampling of deaths leads to overestimation of mortality (underestimation of *e60*): This is a trivial (well-known) consequence. More seriously, it distorts the balance between categories (`sex`, `region`) in unpredictable ways.

# AFT regression

The OLS model is an example of what is called  an *accelerated failure time (AFT)* model in survival analysis, a competitor to the *proportional hazards (PH)* model. 
A common assumption about old age mortality is that a *Gompertz* distribution fits well.

So we go back to the full data situation and fit a Gompertz AFT model. The *survreg* function does not include the Gompertz distribution, but the *aftreg* function in the package **eha** does.

```{r aft, echo = FALSE}
fit.g <- aftreg(Surv(exit, event) ~ sex, data = ols, dist = "gompertz", param = "lifeExp")
summary(fit.g)
```

```{r plotln}
plot(fit.g, fn = "haz", xlab = "Years after 60.")
```
